The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-04 rockerverse.Rmd
fix some widows and orphans
https://en.wikipedia.org/wiki/Widows_and_orphans

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

@@

**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018} for dynamic documents \citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
\pkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g. demonstrated in the DockFlow project (https://dockflow.org).
It introduces new fields to the R Markdown document header, allowing users to manually declare the dependencies, including versions, for rendering the document.
The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
An RStudio addin even allows compilation of documents with the single push of a button.

@@

Containers can also serve as useful playgrounds to create environments ad hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.

First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
The container does not affect their regular system.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occurring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

@@

The prevalence of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commercial products are publicly available.

@@

The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
For example, one model quickly identifies whether a message is from a customer \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a machine learning model and exposes it through a \pkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker container.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

This article is a snapshot of the R corner in a universe of applications built with a many-faced piece of software, Docker.
`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
Docker has become synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments as it can easily move processing between machines (e.g., locally, a cloud provider's VM, another cloud provider's Container-as-a-Service).
Reproducible environments, scalability and efficiency, and portability across platforms/infrastructures are the common themes behind R packages, use cases, and applications in this work.

The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
But the applications also point to the challenges of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
Instructions can be outsourced into distributable scripts and then copied into the image during build, but that makes `Dockerfile`s harder to read.
Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
Despite the different image stacks presented here, the pervasiveness of Rocker images can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker Project's suite of images, more systems will be built successfully, but they may also be built behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

@@

The ordering of authors following DN and DE is alphabetical.
DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DE wrote the introduction and the section about containerisation and the Rocker Project, and reviewed all sections.
DB wrote the section on \pkg{outsider}.
GD contributed the CARD.com use case.
RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
LK contributed content on \pkg{dockr}.
SL contributed content on RStudio's usage of Docker.
BM wrote the section on research compendia and made the project Binder ready.
HN \& JN co-wrote the section on the T-Mobile use case.
KR wrote the section about \pkg{holepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS templates.
TS \& JW wrote the section on CyVerse.
CvP wrote the section on the usage of Docker containers in Dodona.
CW wrote the sections on Whole Tale and contributed content about publication reproducibility audits.
NX contributed content on \pkg{liftr}.
All authors approved the final version.
This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

>>>>

Docker is not the only containerisation software.
**Singularity** stems from the domain of high-perfomance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

The _Bioconductor_ `Dockerfile` selects the desired R version from Rocker images, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
A strength of this approach is that the responsibility for complex software configuration and customization is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is the `bioconductor_docker:devel` image, which emulates the _Bioconductor_ environment for nightly builds as closely as possible.
This image contains the environment variables and the system dependencies needed to install and check almost all _Bioconductor_ software packages (1813 out of 1823).
It saves users and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

@@

**\CRANpkg{liftr}** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018}.
The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
\pkg{liftr} makes the dynamic R Markdown document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g. demonstrated in the DockFlow project (https://dockflow.org).
It introduces new fields to the document header, allowing users to manually declare the versioned dependencies required for rendering the document.
The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
An RStudio addin even allows compilation of documents with the single push of a button.

@@

Containers can also serve as playgrounds and provide specific or ad hoc environments for the purposes of developing R packages.
These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.

First, such containers can greatly facilitate **fixing bugs and code evaluation**, because developers and users can readily start a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
The container can later be discarded and does not affect their regular system.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation without writing a `Dockerfile`.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occurring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

@@

The prevalence of Docker in industry naturally leads to the use of R in containers, as companies already manage platforms in Docker containers.
These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commercial products are publicly available.

@@

The AI @ T-Mobile team created a set of machine learning models for natural language processing to help customer care agents manage text-based messages from customers \citep{t-mobile_enterprise_2018}.
For example, one model identifies whether a message is from a customer \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a such a model and exposes it through a \pkg{plumber} API, then someone else on the marketing team can write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker container.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

This article is a snapshot of the R corner in a universe of applications built with a many-faced piece of software, Docker.
`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as developers and IT operators, and between participants in the communication of knowledge, such as researchers or students.
Docker has become synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments because it can easily move processing between machines, e.g., locally, a cloud provider's VM, another cloud provider's Container-as-a-Service.
Reproducible environments, scalability \& efficiency, and portability across infrastructures are the common themes behind R packages, use cases, and applications in this work.

The projects presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
But the applications also point to the challenges of keeping up with a continuously evolving landscape.
Some use cases have considerable overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems with Docker to serve one's specific needs, such as an independent Docker image stack, leads to parallel developments.
This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
Instructions can be outsourced into distributable scripts and then copied into the image during build, but that makes `Dockerfile`s harder to read.
Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
Despite the different image stacks presented here, the pervasiveness of Rocker images can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more working systems will be built, but they may simply work behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

@@

The ordering of authors following DN and DE is alphabetical.
DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DE wrote the introduction and the section about containerisation and the Rocker Project, and reviewed all sections.
DB wrote the section on \pkg{outsider}.
GD contributed the CARD.com use case.
RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
LK contributed content on \pkg{dockr}.
SL contributed content on RStudio's usage of Docker.
BM wrote the section on research compendia and made the project Binder ready.
HN \& JN co-wrote the section on the T-Mobile use case.
KR wrote the section about \pkg{holepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS templates.
TS \& JW wrote the section on CyVerse.
CvP wrote the section on the usage of Docker containers in Dodona.
CW wrote the sections on Whole Tale and contributed content about publication reproducibility audits.
NX contributed content on \pkg{liftr}.
All authors approved the final version.
This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide details on the respective contributions.
